---
title: "DATA 606 Final Project"
author: "Samuel I Kigamba"
date: "December 9, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r echo=FALSE}
# load data

```


### Part 1 - Introduction


### Part 2 - Data


### Part 3 - Exploratory data analysis


### Part 4 - Inference


### Part 5 - Conclusion


### References


### Appendix (optional)

Remove this section if you don't have an appendix



## TEST: Population vs Crime Ratio

In this section we will create a linear regression model to see if there exists a strong relationship between population and Crime Ratio.

### Create a function to calculate the correlation between population and Crime Ratio and round it to 4 decimal digits

```{r, message=FALSE, warning=FALSE}
findCorrelation <- function() {
  x = crime_data1$Population
  y = crime_data1$Crime_ratio
  corr = round(cor(x, y),4)
  print (paste0("Correlation = ",corr))
  return (corr)
}

c = findCorrelation()

```

### Create a function for Linear Model for Population vs Crime_ratio

```{r, message=FALSE, warning=FALSE}

findStatsFunction <- function() {
  m = lm (Population ~ Crime_ratio, data = crime_data1)
  s = summary(m)
  print(s)
  
  slp = round(m$coefficients[2], 4)
  int = round(m$coefficients[1], 4)

  return (m)
}
m = findStatsFunction()

```

\[ Population^ = 198426094 + (3596492 ??? Crime_ratio) \]


### Display the Linear Model

```{r, message=FALSE, warning=FALSE}
plot = ggplot(crime_data1, aes(Crime_ratio, Population, group = 1)) +
  geom_point(colour="blue") +
  xlab("Total Crime") +
  ylab("Population Size") +
  labs(title = "Population Size vs Crime Ratio") +
  geom_abline(aes(slope=round(m$coefficients[2], 4), intercept=round(m$coefficients[1], 4), color="red"))

print(plot)

```

### Regression Statistics

Linear Regression Equation:

\[ Population^ = 198426094 + (3596492 ??? Crime_ratio) \]

Multiple R-Square:0.9352

R-Square:0.9323

Description:Strong correlation Model fits the data.


Hypothesis Testing

H_0 : 	 Null Hypothesis 	 There is no relationship between Crime Ratio and population.
H_A : 	 Alternative Hypothesis 	 There is a relationship between Crime Ratio and population.

Here the multiple R value is 0.9352 which shows that there is significant correlation between Crime Ratio and population. Also the value of R square is 0.9352 which shows the extent to which the Crime Ratio affects the population. Therefore, we reject the null hypothesis (H_0) and accept the Alternative hypothesis (H_1).
